---
title: "Report on Gun Murders"
author: "John Jordan"
date: "12 February 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a report on 2010 gun murder rates obtained from FBI reports. The original data was obtained from [this Wikipedia page](https://en.wikipedia.org/wiki/Murder_in_the_United_States_by_state). The work for this was done based on different parts of different courses from the HarvardX Data Science Professional Certificate Program. 

Will use the tidyverse library plus ggthemes and ggrepel packages:

```{r loading-libs, message=FALSE}
library(tidyverse)
library(ggthemes)
library(ggrepel)
```
  
and load the data already wrangled:

```{r}
load("rda/murders.rda")
```

## Murder rate by state (barplot)

Can note the large state to state variability by generating a barplot showing the murder rate by state: 

*Murder rate calculated by total number of murders divided by 100,000

```{r murder-rate-by-state, echo=FALSE}
murders %>% mutate(abb = reorder(abb, rate)) %>%
  ggplot(aes(abb, rate)) +
  geom_bar(width = 0.5, stat = "identity", color = "black") +
  coord_flip() +
  xlab("State Abbreviation") +
  ylab("Murder Rate") +
  ggtitle("Murder Rate by State") +
  theme_economist()
```

## Murder rate by state (scatterplot)
Based on total number of murders and colour-coded by region

Plot also clearly shows that states with higher populations also have more murders. 

```{r scatterplot, echo=FALSE}
national_murder_rate <- murders %>% summarise(rate = sum(total) / sum(population) * 10^6) %>% .$rate

murders %>% ggplot(aes(population/10^6, total, label = abb)) +
  geom_abline(intercept = log10(national_murder_rate), lty = 2, color = "darkgrey") + 
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Population in millions (log scale)") +
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region") +
  theme_economist()
```

## Murder rate by region (boxplot)
Examine the murder rate based on region. Results clearly show that the South has a higher murder rate than other regions. 

Individual points for each state.

```{r regional boxplot, echo=FALSE}
murders %>% ggplot(aes(y=rate, x=region, fill = region)) + 
  geom_boxplot(outlier.colour = "black") + 
  geom_point() +
  xlab("Murder Rate") + 
  ylab("Region") +
  ggtitle("Murder Rate by Region") +
  theme_economist()
