---
title: "Report on Gun Murders"
author: "John Jordan"
date: "12 February 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a report on 2010 gun murder rates obtained from FBI reports. The original data was obtained from [this Wikipedia page](https://en.wikipedia.org/wiki/Murder_in_the_United_States_by_state). The work for this was done based on different parts of different courses from the HarvardX Data Science Professional Certificate Program. 

Will use the tidyverse library:

```{r loading-libs, message=FALSE}
library(tidyverse)
```
  
and load the data already wrangled:

```{r}
load("rda/murders.rda")
```

## Murder rate by state (barplot)

We note the large state to state variability by generating a barplot showing the murder rate by state:

```{r murder-rate-by-state, echo=FALSE}
murders %>% mutate(abb = reorder(abb, rate)) %>%
  ggplot(aes(abb, rate)) +
  geom_bar(width = 0.5, stat = "identity", color = "black") +
  coord_flip() +
  xlab("State Abbreviation") +
  ylab("Murder Rate")
```

## Murder rate by state (scatterplot)

Based on total number of murders and colour-coded by region

```{r scatterplot, echo=FALSE}
library(ggthemes)
library(ggrepel)
r <- murders %>% summarise(rate = sum(total) / sum(population) * 10^6) %>% .$rate

murders %>% ggplot(aes(population/10^6, total, label = abb)) +
  geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") + 
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Population in millions (log scale)") +
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region") +
  theme_economist()
```

